---
title: "Factors associated with opportunities for improvement in trauma care"
output: bookdown::pdf_document2
bibliography: citation.bib
always_allow_html: true
csl: bmcemerg.csl
link-citations: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```
```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```
\newpage

```{r library, include=FALSE}
library(readr)
library(naniar)
library(plyr)
library(dplyr)
library("questionr")
library(table1)
library(ggplot2)
library(gridExtra)
library(gtsummary)
library(DiagrammeR)
library(PRISMAstatement)
library(knitr)
library(gt)
library(tidyverse)
library(glue)
library(tinytex)
library(rmarkdown)
library(RMariaDB)
library(DBI)
library(keyring)
library(dotenv)
library(gtsummary)
library(flextable)
library(kableExtra)
library(ggpubr)
library(lubridate)
library(parameters)
library(rofi)
```



```{r include=FALSE}
conn <- DBI::dbConnect(drv = RMariaDB::MariaDB(),
                       user = 'husseina',
                       password = Sys.getenv("DB_PASSWORD"),
                       db = "opportunities_for_improvement")


dataset.names <- setNames(nm = c("swetrau", "fmp", "atgarder", "problem"))
datasets <- lapply(dataset.names, function(dataset.name) dbReadTable(conn = conn, name = dataset.name))

attach(datasets)

swetrau$arrival <- as.POSIXct(strptime(swetrau$DateTime_ArrivalAtHospital, format = "%Y-%m-%d %H:%M"))
fmp$arrival <- as.POSIXct(strptime(fmp$Ankomst_te, format = "%Y%m%d %H:%M"))
problem$arrival <- as.POSIXct(strptime(problem$Ankomst_te, format = "%Y%m%d %H:%M"))
swetrau$id <- paste(swetrau$arrival, swetrau$PersonIdentity, swetrau$TempIdentity)
fmp$id <- paste(fmp$arrival, fmp$Personnummer, fmp$Reservnummer)
problem$id <- paste(problem$arrival, problem$Personnummer, problem$Reservnummer)

# Combine datasets
combined.data <- merge(fmp, problem, by = "id", all.x = TRUE)
combined.data <- merge(combined.data, swetrau, by = "id", all.x = TRUE)

detach(datasets)
combined.datasets<-filter(combined.data, as.Date(strptime(combined.data$Ankomst_te.x, format = "%Y%m%d %H:%M")) >="2017-01-01")

combined.datasets$ofi <- create_ofi(combined.datasets)
combined.datasets$ofi <- with(combined.datasets, ifelse(is.na(ofi) | ofi == "No", "No", "Yes"))


# Gör en kolumn med intub: 1, intub ED, 2. Ej intub. 3. Intub prehosp.
combined.datasets$intub <- with(combined.datasets, ifelse(`pre_intubated` == 1 & is.na(combined.datasets$pre_intubated) == FALSE, 3, `ed_intubated`))
# Förklaring: tub<- om pre inte är Na och är 1(ja), sant: konvertera till 3, Falskt: använd ed_intubated (1 är ja och 2 är nej)

# Gör om 999 till Na
ab<-combined.datasets %>% replace_with_na(replace = list( "res_survival" = 999))
ac<-ab %>% replace_with_na(replace = list( "intub" = 999))
ada<-ac %>% replace_with_na(replace = list( "host_care_level" = 999))
adab<-ada %>% replace_with_na(replace = list( "ed_sbp_rtscat" = 999))
adac<-adab %>% replace_with_na(replace = list( "ed_rr_rtscat" = 999))
adad<-adac %>% replace_with_na(replace = list( "pre_rr_rtscat" = 999))
adae<-adad %>% replace_with_na(replace = list( "ed_sbp_value" = 999))
adah<-adae %>% replace_with_na(replace = list( "ed_rr_value" = 999))
adag<-adah %>% replace_with_na(replace = list( "pre_rr_value" = 999))
adaj<-adag %>% replace_with_na(replace = list( "ed_gcs_sum" = 999))
swetrau<-adaj %>% replace_with_na(replace = list( "pre_gcs_sum" = 999))

# Gör om blodtryck till RTS
swetrau$ed_sbp_value[is.na(swetrau$ed_sbp_value)]<-350

swetrau$ed_sbp_value[swetrau$ed_sbp_value>=1 & swetrau$ed_sbp_value<=49]<-1
swetrau$ed_sbp_value[swetrau$ed_sbp_value>=50 & swetrau$ed_sbp_value<=75]<-2
swetrau$ed_sbp_value[swetrau$ed_sbp_value>=76 & swetrau$ed_sbp_value<=89]<-3
swetrau$ed_sbp_value[swetrau$ed_sbp_value>89 & swetrau$ed_sbp_value<=298]<-4
swetrau$ed_sbp_value[swetrau$ed_sbp_value==350]<-5

# Tar RTS värdet från ed, om det inte finns så använder den vanliga blodtrycket, som är grupperat enligt RTS.
swetrau$sbp_rts<-with(swetrau, ifelse(is.na(swetrau$ed_sbp_rtscat)==FALSE,ed_sbp_rtscat,ed_sbp_value))

# Samma med ed_RR
swetrau$ed_rr_value[is.na(swetrau$ed_rr_value)]<-350

swetrau$ed_rr_value[swetrau$ed_rr_value>=1 & swetrau$ed_rr_value<=5]<-1
swetrau$ed_rr_value[swetrau$ed_rr_value>=6 & swetrau$ed_rr_value<=9]<-2
swetrau$ed_rr_value[swetrau$ed_rr_value>=10 & swetrau$ed_rr_value<=29]<-4
swetrau$ed_rr_value[swetrau$ed_rr_value>29 & swetrau$ed_rr_value<=70]<-3
swetrau$ed_rr_value[swetrau$ed_rr_value==99]<-5
swetrau$ed_rr_value[swetrau$ed_rr_value==350]<-5

swetrau$ed_rr_rts<-with(swetrau, ifelse(is.na(swetrau$ed_rr_rtscat)==FALSE,ed_rr_rtscat,ed_rr_value))

# Samma med pre_RR
swetrau$pre_rr_value[swetrau$pre_rr_value>=1 & swetrau$pre_rr_value<=5]<-1
swetrau$pre_rr_value[swetrau$pre_rr_value>=6 & swetrau$pre_rr_value<=9]<-2
swetrau$pre_rr_value[swetrau$pre_rr_value>=10 & swetrau$pre_rr_value<=29]<-4
swetrau$pre_rr_value[swetrau$pre_rr_value>29 & swetrau$pre_rr_value<=70]<-3
swetrau$pre_rr_value[swetrau$pre_rr_value==99]<-5
swetrau$pre_rr_value[swetrau$pre_rr_value==350]<-5

swetrau$pre_rr_rts<-with(swetrau, ifelse(is.na(swetrau$pre_rr_rtscat)==FALSE ,pre_rr_rtscat,pre_rr_value))

swetrau$rr_rts<-with(swetrau, ifelse(intub == 3 & pre_rr_rts !=5,pre_rr_rts,ed_rr_rts))

swetrau$ed_gcs_sum[is.na(swetrau$ed_gcs_sum)]<-350

# Konvertera GCS till mild 13-15 (3), moderate 9-12 (2), severe 3-8. (1). 5 = MISSING
swetrau$ed_gcs_sum[swetrau$ed_gcs_sum>=3 & swetrau$ed_gcs_sum<=8]<-1
swetrau$ed_gcs_sum[swetrau$ed_gcs_sum>=9 & swetrau$ed_gcs_sum<=12]<-2
swetrau$ed_gcs_sum[swetrau$ed_gcs_sum>=13 & swetrau$ed_gcs_sum<=15]<-3
swetrau$ed_gcs_sum[swetrau$ed_gcs_sum==99]<-5
swetrau$ed_gcs_sum[swetrau$ed_gcs_sum==350]<-5

# Gör samma med pre_GCS
swetrau$pre_gcs_sum[is.na(swetrau$pre_gcs_sum)]<-350

swetrau$pre_gcs_sum[swetrau$pre_gcs_sum>=3 & swetrau$pre_gcs_sum<=8]<-1
swetrau$pre_gcs_sum[swetrau$pre_gcs_sum>=9 & swetrau$pre_gcs_sum<=12]<-2
swetrau$pre_gcs_sum[swetrau$pre_gcs_sum>=13 & swetrau$pre_gcs_sum<=15]<-3
swetrau$pre_gcs_sum[swetrau$pre_gcs_sum==99]<-5
swetrau$pre_gcs_sum[swetrau$pre_gcs_sum==350]<-5

# Skapa gcs_sum som blir vår nya gcs, 5=missing
swetrau$gc<-with(swetrau, ifelse(intub == 3 & pre_gcs_sum !=5,pre_gcs_sum,ed_gcs_sum))
swetrau$gcs<-with(swetrau, ifelse(intub == 1 & gc ==5,99,gc))
swetrau$gcs_sum<-with(swetrau, ifelse(intub == 3 & gcs ==5,99,gcs))

# Egen datafram för räkna missing data
fake.var<-c("Gender","res_survival","intub","host_care_level","rr_rts","sbp_rts","gcs_sum")
rak.var<-c("id","ofi","Gender","res_survival","intub","host_care_level","rr_rts","sbp_rts","gcs_sum")
cont.var<-c("ISS","dt_ed_first_ct","pt_age_yrs")
cont.procedure.var<-c("ISS","dt_ed_first_ct","pt_age_yrs","dt_ed_emerg_proc")
cat.var<-c("ofi","Gender","res_survival","intub","host_care_level","rr_rts","sbp_rts","gcs_sum")

# Gör ny dataframe för att räkna missing data
ba<-select(swetrau,rak.var,cont.var)
kl<-select(swetrau,fake.var,cont.var)
na.kl<-na.omit(kl)

# Sätt ihop databaser
calc<-ba

# Gör om till Na
calh<-calc %>% replace_with_na(replace = list('pt_age_yrs'=11))
calj<-calh %>% replace_with_na(replace = list('pt_age_yrs'=13))
calk<-calj %>% replace_with_na(replace = list('pt_age_yrs'=14))

# Gör om missing i vitalparametrar för att beräkna missing
calb<-calk %>% replace_with_na(replace = list('rr_rts'=5))
cald<-calb %>% replace_with_na(replace = list('sbp_rts'=5))
cale<-cald %>% replace_with_na(replace = list('gcs_sum'=5))

sew<-as.data.frame(freq.na(cale))

# Välj variabler
nks<-select(calk,cat.var,cont.var)

# Ta bort alla missing data
ki<- na.omit(nks)

# Kopiera dataframe
ae<-ki
aa<-ki

# Gör om till faktorer
str(aa)
aa$ofi<-as.factor(aa$ofi)
aa$Gender<-as.factor(aa$Gender)
aa$host_care_level<-as.factor(aa$host_care_level)
aa$intub<-as.factor(aa$intub)
aa$res_survival<-as.factor(aa$res_survival)
aa$rr_rts<-as.factor(aa$rr_rts)
aa$sbp_rts<-as.factor(aa$sbp_rts)
aa$gcs_sum<-as.factor(aa$gcs_sum)
aa$ISS<-as.numeric(aa$ISS)
aa$dt_ed_first_ct<-as.numeric(aa$dt_ed_first_ct)
aa$pt_age_yrs<-as.numeric(aa$pt_age_yrs)

# Ritar ut alla värden på graf
pairs(aa,col=aa$ofi)

# Resultat för varje variabel, unadjusted

sex<-glm(ofi~Gender,data = aa,family = 'binomial')
summary(sex)

surv<-glm(ofi~res_survival,data = aa,family = 'binomial')
summary(surv)

tub<-glm(ofi~intub,data = aa,family = 'binomial')
summary(tub)

care<-glm(ofi~host_care_level,data = aa,family = 'binomial')
summary(care)

resp<-glm(ofi~rr_rts,data = aa,family = 'binomial')
summary(resp)

sysbp<-glm(ofi~sbp_rts,data = aa,family = 'binomial')
summary(sysbp)

gcs_var<-glm(ofi~gcs_sum,data = aa,family = 'binomial')
summary(gcs_var)

injury<-glm(ofi~ISS,data = aa,family = 'binomial')
summary(injury)

first_ct<-glm(ofi~dt_ed_first_ct,data = aa,family = 'binomial')
summary(first_ct)


age<-glm(ofi~pt_age_yrs,data = aa,family = 'binomial')
summary(age)


# Table 1
aa$ofi <-
  factor(aa$ofi,
         levels=c("No","Yes"),
         labels=c("No opportunity for improvement",
                  "Opportunity for improvement"))

aa$Gender <-
  factor(aa$Gender, levels=c("M","K"),
         labels=c("Male",
                  "Female"))

aa$res_survival <-
  factor(aa$res_survival, levels=c("2","1"),
         labels=c("Alive",
                  "Dead"))

aa$intub <-
  factor(aa$intub, levels=c("2","3","1"),
         labels=c("None","Pre-intub",
                  "ED-intub"))

aa$host_care_level <-
  factor(aa$host_care_level, levels=c("1","2","3","4","5"),
         labels=c("ED","Admissioned","Surgical ward","Specialized ward","ICU"))

aa$sbp_rts <-
  factor(aa$sbp_rts, levels=c("4","3","2","1","0","5"),
         labels=c(">89","76-89","50-75","1-49","0","Missing"))

aa$gcs_sum <-
  factor(aa$gcs_sum, levels=c("3","2","1","5","99"),
         labels=c("13-15","9-12","3-8","Missing","Pre-intub"))

aa$rr_rts <-
  factor(aa$rr_rts, levels=c("4","3","2","1","0","5"),
         labels=c("10-29",">29","6-9","1-5","0","Missing"))

label(aa$ofi)      <- "Outcome"
label(aa$ISS)      <- "Injury Severity Score"
label(aa$Gender)       <- "Sex"
label(aa$intub)     <- "Intubation"
label(aa$pt_age_yrs)       <- "Age"
label(aa$res_survival)       <- "Survival after 30 days"
label(aa$host_care_level)       <- "Highest hospital care level"
label(aa$intub)       <- "Intubation"
label(aa$rr_rts)       <- "Respiratory rate"
label(aa$sbp_rts)       <- "Systolic Blood Pressure"
label(aa$gcs_sum)       <- "GCS"
label(aa$dt_ed_first_ct)       <- "Time to first CT"

units(aa$pt_age_yrs)       <- "Years"
units(aa$dt_ed_first_ct)   <- "minutes"
units(aa$sbp_rts)       <- "RTS"
units(aa$rr_rts)       <- "RTS"
units(aa$gcs_sum)       <- "RTS"

# Resultat för adjusted variabels
outcome_log<-glm(ofi~.,data = aa,family = 'binomial')
summary(outcome_log)

opo<- as.data.frame(table1(~ Gender + pt_age_yrs + res_survival + host_care_level + intub + ISS + rr_rts + sbp_rts +
         gcs_sum + dt_ed_first_ct | ofi, data=aa))

# Kod för resultat-delen
# Värden för stapel-diagram
calculation<-aa
calculation$host_care_level <-
  factor(calculation$host_care_level, levels=c("ED","Admissioned","surgical ward","Specialized ward","ICU"),
         labels=c("ED","Adm","Surg","Spec","ICU"))

calculation$intub <-
  factor(calculation$intub, levels=c("Pre-intub","ED-Intub","None"),
         labels=c("Pre-Intub", "Intub","Not Intub"))

calculation$gcs_sum <-
  factor(calculation$gcs_sum, levels=c("Mild: 13-15","Moderate: 9-12","Severe: 3-8","Missing","Pre-intubated"),
         labels=c("13-15","9-12","3-8","Missing","Pre-intub"))

```

```{r include=FALSE, warning=FALSE}
app.ba<-select(swetrau,rak.var,cont.procedure.var)
app.kl<-select(swetrau,fake.var,cont.procedure.var)
app.calc<-app.ba

app.calh<-app.calc %>% replace_with_na(replace = list('pt_age_yrs'=11))
app.calj<-app.calh %>% replace_with_na(replace = list('pt_age_yrs'=13))
app.calk<-app.calj %>% replace_with_na(replace = list('pt_age_yrs'=14))

# Gör om missing i vitalparametrar för att beräkna missing
app.calb<-app.calk %>% replace_with_na(replace = list('rr_rts'=5))
app.cald<-app.calb %>% replace_with_na(replace = list('sbp_rts'=5))
app.cale<-app.cald %>% replace_with_na(replace = list('gcs_sum'=5))

app.sew<-as.data.frame(freq.na(app.cale))

app.nks<-select(app.calk,cat.var,cont.procedure.var)

app.ki<- na.omit(app.nks)
app.ae<-app.ki
app.aa<-app.ki

app.aa$ofi<-as.factor(app.aa$ofi)
app.aa$Gender<-as.factor(app.aa$Gender)
app.aa$host_care_level<-as.factor(app.aa$host_care_level)
app.aa$intub<-as.factor(app.aa$intub)
app.aa$res_survival<-as.factor(app.aa$res_survival)
app.aa$rr_rts<-as.factor(app.aa$rr_rts)
app.aa$sbp_rts<-as.factor(app.aa$sbp_rts)
app.aa$gcs_sum<-as.factor(app.aa$gcs_sum)
app.aa$ISS<-as.numeric(app.aa$ISS)
app.aa$dt_ed_first_ct<-as.numeric(app.aa$dt_ed_first_ct)
app.aa$dt_ed_emerg_proc<-as.numeric(app.aa$dt_ed_emerg_proc)
app.aa$pt_age_yrs<-as.numeric(app.aa$pt_age_yrs)
str(app.aa)

app.aa$ofi <-
  factor(app.aa$ofi,
         levels=c("No","Yes"),
         labels=c("No opportunity for improvement",
                  "Opportunity for improvement"))

app.aa$Gender <-
  factor(app.aa$Gender, levels=c("M","K"),
         labels=c("Male",
                  "Female"))

app.aa$res_survival <-
  factor(app.aa$res_survival, levels=c("1","2"),
         labels=c("Dead",
                  "Alive"))

app.aa$intub <-
  factor(app.aa$intub, levels=c("3","1","2"),
         labels=c("Pre-hospital Intubation",
                  "ED Intubation","Not Intubated"))

app.aa$host_care_level <-
  factor(app.aa$host_care_level, levels=c("1","2","3","4","5"),
         labels=c("ED","Admissioned","surgical ward","Specialized ward","ICU"))

app.aa$sbp_rts <-
  factor(app.aa$sbp_rts, levels=c("4","3","2","1","0","5"),
         labels=c(">89","76-89","50-75","1-49","0","Missing"))

app.aa$gcs_sum <-
  factor(app.aa$gcs_sum, levels=c("3","2","1","5","99"),
         labels=c("Mild: 13-15","Moderate: 9-12","Severe: 3-8","Missing","Pre-intubated"))

app.aa$rr_rts <-
  factor(app.aa$rr_rts, levels=c("3","4","2","1","0","5"),
         labels=c(">29","10-29","6-9","1-5","0","Missing"))

label(app.aa$ofi)      <- "Outcome"
label(app.aa$ISS)      <- "Injury Severity Score"
label(app.aa$Gender)       <- "Sex"
label(app.aa$intub)     <- "Intubation"
label(app.aa$pt_age_yrs)       <- "Age"
label(app.aa$res_survival)       <- "Survival after 30 days"
label(app.aa$host_care_level)       <- "Highest hospital care level"
label(app.aa$intub)       <- "Intubation"
label(app.aa$rr_rts)       <- "Respiratory rate"
label(app.aa$sbp_rts)       <- "Systolic Blood Pressure"
label(app.aa$gcs_sum)       <- "GCS"
label(app.aa$dt_ed_first_ct)       <- "Time to first CT"
label(app.aa$dt_ed_emerg_proc)       <- "Time to intervention"

units(app.aa$dt_ed_emerg_proc)   <- "minutes"
units(app.aa$pt_age_yrs)       <- "Years"
units(app.aa$dt_ed_first_ct)   <- "minutes"
units(app.aa$sbp_rts)       <- "RTS"
units(app.aa$rr_rts)       <- "RTS"

app.outcome_log<-glm(ofi~.,data = app.aa,family = 'binomial')
summary(app.outcome_log)

app.log<-tbl_regression(app.outcome_log,exponentiate = TRUE)

app.log
```


```{r, include=FALSE}

DiagrammeR::grViz("
digraph graph2 {

graph [layout = dot, rankdir = RR]

# node definitions with substituted label text
node [shape = rectangle, width = 4, fillcolor = Biege]
a [label = '@@1']
b [label = '@@2']
c [label = '@@3']
d [label = '@@4']
e [label = '@@5']
f [label = '@@6']
g [label = '@@7']

a -> b -> c {rank = same; b;g;}
c -> d [label='Audit filters identified/ OFI are likely']
c -> e [label='Second opinion needed'] {rank = same; c; e;}
e -> g
d -> f 
e -> d
c -> g [label='No audit filters identified/ OFI are unlikely']

}

graph <- select_nodes(graph)
graph <-
  graph %>%
  nudge_node_positions_ws(
    dx = 0, dy = 10)

[1]: paste0('The occurence of a trauma')
[2]: paste0('Registration in SweTrau')
[3]: paste0('Primary review of audit filters by a nurse')
[4]: paste0('Multidisciplinary morbidity and mortality conference')
[5]: paste0('Secondary review by a colleague')
[6]: paste0('Consensus regarding existance of OFI')
[7]: paste0('Exclusion')


") %>% DiagrammeRsvg::export_svg() %>% charToRaw() %>% rsvg::rsvg_pdf("flowchart_process.pdf")
```

```{r echo=F,results='asis',error=F,warning=F}
sewe<-filter(sew, missing != 0)

rownames(sewe)[rownames(sewe) == "res_survival"] = "Survival after 30 days"
rownames(sewe)[rownames(sewe) == "gcs_sum"] = "GCS"
rownames(sewe)[rownames(sewe) == "rr_rts"] = "Respiratory rate"
rownames(sewe)[rownames(sewe) == "intub"] = "Intubation"
rownames(sewe)[rownames(sewe) == "host_care_level"] = "Highest hospital care level"
rownames(sewe)[rownames(sewe) == "ofi"] = "Outcome"
rownames(sewe)[rownames(sewe) == "pt_age_yrs"] = "Age"
rownames(sewe)[rownames(sewe) == "dt_ed_first_ct"] = "Time to first CT"
rownames(sewe)[rownames(sewe) == "sbp_rts"] = "Systolic blood pressure"
rownames(sewe)[rownames(sewe) == "dt_ed_emerg_proc"] = "Time to intervention"
rownames(sewe)[rownames(sewe) == "Gender"] = "Sex"
```

### Factors associated with opportunities for improvement in trauma care: A registry-based retrospective cohort study

<!-- Is this the title? Put as title: in the yaml header. Also, I think you can remove retrospective cohort from the study, "A registry-based study" is enough -->

*Background*: Trauma is one of the leading causes of morbidity and mortality worldwide. Morbidity and mortality review is used to improve the quality of trauma care. During this review specific cases are reviewed to identify opportunities for improvement (OFI). The selection of cases for review is often inefficient. The aim of this study was to assess how certain patient level factors are associated with OFI in trauma care.

*Methods*: We used data from the Karolinska University Hospital (KUH) trauma registry (TrR) and the trauma care quality database (TrcQD) between 2017 and 2021. Our outcome was OFI as defined by the morbidity and mortality review at KUH. We used bi- and multivariable logistic regression to assess the associations between the following patient level factors and OFI: age, sex, respiratory rate, systolic blood pressure, Glasgow Coma Scale, Injury Severity Score, survival at 30 days, highest hospital care level, intubation status and time to CT and intervention. 

*Results*: OFI was identified in 276 (5.94%) out of 4643 patients. The median age for patients with OFI was 49 compared to 42 in patients without OFI. Age, highest hospital care level, injury severity score (ISS), respiratory rate, systolic blood pressure, Glasgow Coma Scale (GCS), survival after 30 days and intubation were statistically significantly associated with OFI. 

*Conclusion*: Higher hospital care level, age and ISS were found to be significantly associated with OFI in trauma care<!-- You list more variables in Results just above, so it looks odd that only these three are mentioned here. Try to summarise your key results here instead, for example "Several patient level factors are associated with OFI"-->. Future research should focus on identifying reasons to why these factors are associated with higher odds of OFI.

\newpage

# Introduction
Trauma is one of the leading causes of mortality and morbidity in all age groups [@David2021;@Ono2021]. There are approximately 4.5 million global deaths each year due to trauma [@James2020], and it is one of the top contributors to disease burden worldwide, measured by disability-adjusted life years [@Murray1996;@Haagsma2015]. Trauma is also one of the most prevalent reasons for intensive care unit admission [@Prin2016], and is therefore resource-intensive.

Patient outcomes are dependent on the quality of their care [@Dogrul2020;@Ghorbani2014;@Kahl2013]. The gold standard method for trauma care quality improvement is morbidity and mortality review, during which specific cases are reviewed to investigate the reasons for morbidity and mortality  [@WHO2009]. This review often aims to establish if deaths were preventable and if there are opportunities for improvement in the care provided. 

Preventable death rates range between 4 and 60% across different settings and countries  [@Zafarghandi2003;@Konadu2020; @Ghorbani2018]. The most common opportunities for improvement relate to airway management...<!-- Add more here, from the sources you already cite plus for example:
** OReilly2013                 :trauma:opportunities:for:improvement:quality:
@article{OReilly2013,
  title =	 {Opportunities for improvement in the management of
patients who die from haemorrhage after trauma},
  volume =	 {100},
  ISSN =	 {1365-2168},
  url =		 {http://dx.doi.org/10.1002/bjs.9096},
  DOI =		 {10.1002/bjs.9096},
  number =	 {6},
  journal =	 {British Journal of Surgery},
  publisher =	 {Oxford University Press (OUP)},
  author =	 {O’Reilly, D and Mahendran, K and West, A and
Shirley, P and Walsh, M and Tai, N},
  year =	 {2013},
  month =	 {Mar},
  pages =	 {749–755}
}

@article{Roy2017,
  title =	 {Learning from 2523 trauma deaths in India-
  opportunities to prevent in-hospital deaths},
  volume =	 {17},
  ISSN =	 {1472-6963},
  url =		 {http://dx.doi.org/10.1186/s12913-017-2085-7},
  DOI =		 {10.1186/s12913-017-2085-7},
  number =	 {1},
  journal =	 {BMC Health Services Research},
  publisher =	 {Springer Science and Business Media LLC},
  author =	 {Roy, Nobhojit and Kizhakke Veetil, Deepa and
  Khajanchi, Monty Uttam and Kumar, Vineet and Solomon, Harris and
  Kamble, Jyoti and Basak, Debojit and Tomson, Göran and von Schreeb,
  Johan},
  year =	 {2017},
  month =	 {Feb},
  file = {[[file:~/projects/bibliography/papers/Roy2017.pdf]]}
}
-->

The morbidity and mortality review process is resource-intensive [@Anderson2021]. Only a small number of cases can be reviewed, so selecting the right cases is crucial. Methods such as audit filters are used to identify cases in which opportunities for improvement are likely. But these methods are fraught with high numbers of false positives and need improvement. The aim of this study was to asses how certain patient level factors are associated with opportunities for improvement in trauma care.

# Methods

### Study design
<!-- Use first person and active voice: we used, we conducted, we selected -->
We used data from the Karolinska University Hospital trauma registry, part of the Swedish National Trauma Registry (SweTrau) [@Swetrau2020], and the local trauma care quality database. Swetrau includes information of pre-hospital, hospital, and post-hospital care in accordance with the Utstein template [@Utstein2009]. Factors such as demographics, vital signs, time to procedure and time to intervention are registered [@Dick1999]. The trauma care quality database includes all trauma patients treated at the Karolinska University Hospital from the year of 2017. This database has the results of the morbidity and mortality review including opportunities for improvement. The trauma registry and trauma care quality database were linked and possible factors associated with opportunities for improvement were extracted. Bi- and multivariable logistic regression were used to determine association with the presence of opportunities for improvement<!-- You need to decide if you want to abbreviate OFI or not. You do so in the abstract, but not in the introduction, and then here again in the methods. I generally recommend avoiding abbreviations as much as possible because it makes the text easier to read-->.

### Setting
In Sweden, pre-hospital care is managed by specialized nurses, emergency medical technicians and specially trained physicians. The trauma patient is triaged by the emergency medical services personnel at the scene. If one or more of the trauma level one criteria are met then the patient is treated as a priority one [@traumalarmskriterier2017]. All priority one trauma patients in Stockholm  are transported to Karolinska University Hospital to receive care by dedicated trauma teams. These teams consist of a trauma surgeon, an anaesthetist, an orthopaedic surgeon, a radiologist and specialized nurses. Karolinska University Hospital in Solna is a level one trauma center, with direct access to radiology, intervention, operation, intensive care and consultants in all associated specialities [@Social2015;@NKS2020].

Fig.\@ref(fig:flowchart-process) describes the morbidity and mortality review process. Specialised nurses review all trauma patients presenting to the Karolinska University Hospital. To facilitate their initial review they apply audit filters to identify cases in which opportunities for improvement are more likely. These filters are systolic blood pressure under 90, GCS less than 9 and not intubated, ISS more than 15 but not admitted to the ICU, time to acute intervention more than 60 minutes, time to computed tomography more than 30 minutes, and death within 30 days after trauma. 

If the specialized nurses decide that there is a potential OFI, the case is selected for inclusion in a morbidity and mortality review conference. During these conferences cases are reviewed by experienced specialists from all the disciplines and professions involved in the trauma team. The presence or absence of opportunities for improvement is a consensus decision among all participants of the conference and is recorded in the trauma care quality database. A plan to solve the problems identified is also presented by the board.

```{r flowchart-process, fig.cap="Flowchart describing the process of trauma cases from arrival until the inclusion in the trauma care quality database", echo=FALSE, fig.align='center'}
knitr::include_graphics("flowchart_process.pdf")
```

### Participants
The trauma registry includes all patients admitted with trauma team activation, regardless of ISS, as well as patients admitted without trauma team activation but found to have an ISS of more than 9. We included all patients present in both the trauma care quality database and trauma registry who had been included in the morbidity and mortality review process between 2017 and 2021. We excluded patients who were younger than 15 years. We also excluded patients with missing covariate data.

### Variables

**Study outcome**
The outcome was the presence of opportunities for improvement, as decided by the morbidity and mortality review board, and defined as a binary variable with the levels "Yes - At least one OFI identified" and "No - No opportunities for improvement identified". Data on this outcome was extracted from the trauma care quality database. 

**Patient level factors**
We selected factors from the trauma registry to be further examined through discussion, partly based on locally used audit filters, standard epidemiological factors and factors frequently registered in Swetrau. We selected the categorical variables sex, survival after 30 days, highest hospital care level, GCS, respiratory rate, systolic blood pressure and if the patient was intubated. We also selected the continuous variables  age, ISS, time from arrival at the hospital until first CT and time from arrival at the hospital until first intervention.

Systolic blood pressure and respiratory rate are registered as either a continuous or a categorical value according to the Revised Trauma Score, but never both. We decided to convert the continuous values, if registered, into categorical values and include a missing category. GCS was also categorized using the Revised Trauma Score levels. We used pre-hospital values for GCS and respiratory rate for patients who were intubated before arrival at the emergency department because these values are not recorded in the emergency department for these patients. 

### Statistical analysis
We use descriptive data to present the baseline characteristics. We used bivariable logistic regression to determine unadjusted association, and multivariable logistic regression to determine adjusted associations with the presence of OFI. We present results as odds ratios (OR) with associated 95% confidence interval. We used a significance level of 5%. We used the programming language R for all analyses [@Rstudio2022]. All statistical analysis were first done on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database to ensure objectivity. We conducted a complete case analysis.

# Results
<!-- In your code it looks like you first combined the datasets, but then you do quite a bit of work in the swetrau dataset. Do you then combine them again? In my opinion it would make more sense to conduct all the work on the combined datasets, instead of jumping back and forth. That would also make the text below easier to follow, if you report all exclusions based on the combined data. So, first combine then exclude --> 
The baseline characteristics are presented in Table \@ref(tab:sample-characteristics). Out of `r nrow(combined.datasets)` patients included in TrR between 2017 and 2021, `r nrow(combined.datasets)-nrow(na.kl)` patients were excluded due to missing data and age under 15, leaving a total of `r nrow(na.kl)` patients eligible for the study. The most prevalent reason for exclusion was missing data in the factor `r rownames(sewe[sewe$missing == sewe[1,1],])` with `r sewe[1,1]` patients, `r round(sewe[1,1]/nrow(combined.datasets)*100, digits = 2)`% of patients registered. The number of patients excluded due to missing data in each factor is shown in Table \@ref(tab:missing-data). In the TrcQD `r nrow(combined.datasets)-sum(is.na(combined.datasets$ofi))` of the patients were applicable for the study. After the combination of data from the TrR and the TrcQD and the removal of unqualified patients, a study group of `r nrow(aa)` is produced. Among the `r nrow(aa)` included patients, `r opo[3,4]` were male. The median age [Min, Max] for all patients was `r opo[7,4]`. The survival rate after 30 days among included patients was `r round(sum(aa$res_survival=='Alive')/nrow(aa)*100, digits = 2)`%.

```{r sample-characteristics, fig.align='center', echo = FALSE}
table1_t <- table1(~ pt_age_yrs + Gender + res_survival + host_care_level + intub + ISS + rr_rts + sbp_rts +
         gcs_sum + dt_ed_first_ct | ofi, data=aa, caption = "Sample characteristics")

table1.conv <- t1kable(table1_t) %>%
  kable_styling(latex_options = "HOLD_position", full_width = F, position = "center")

table1.conv
```

```{r missing-data, fig.cap="Note that the total number of missing values is more than the total number of patients as some patients may have missing values in more than one factor", echo = FALSE, fig.align='center'}
missing_table <- knitr::kable(sewe[,1:2], format = "markdown", caption = "<strong> Missing  values in each factor</strong>")

missing_table
```

### Factors associated with opportunity for improvement
An OFI was identified in `r sum(aa$ofi == "Opportunity for improvement")` (`r round(sum(aa$ofi == "Opportunity for improvement")/nrow(aa)*100, digits = 2)`%) of the patients. The unadjusted and adjusted associations of selected factors with OFI are presented in Table \@ref(tab:var-adjusted). Statistical significance before and after adjustment was identified for age, ISS, GCS and highest hospital care level. The factors respiratory rate, systolic blood pressure and intubation were significantly associatied with OFI only before the adjustment for other factors. The factor survival after 30 days was significantly associated with OFI only after adjustment for other factors. As seen in Table \@ref(tab:sample-characteristics) the odds of OFI in patients with unregistered GCS was 0.2 compared to patients with mild GCS after the adjustment for other factors, which was statistically significant. 

```{r j, include=FALSE}
tbl_merge_test <-
    tbl_merge(tbls = list(aa %>%
                              select(ofi, pt_age_yrs, Gender, res_survival, rr_rts, sbp_rts, gcs_sum, intub, host_care_level, ISS, dt_ed_first_ct) %>%
                              tbl_uvregression(
                                  method = glm,
                                  y = ofi,
                                  method.args = list(family = binomial),
                                  exponentiate = TRUE,
                                  pvalue_fun = ~style_pvalue(.x, digits = 2),
                                  hide_n = TRUE
                              )%>% bold_p(), tbl_regression(outcome_log,exponentiate = TRUE,pvalue_fun = ~style_pvalue(.x, digits = 2)) %>% bold_p() %>% bold_labels()))


hellp<- tbl_merge_test %>% modify_spanning_header(list (estimate_1 ~ "Unadjusted", ci_1 ~ "Unadjusted", p.value_1 ~ "Unadjusted", estimate_2 ~ "Adjusted", ci_2 ~ "Adjusted", p.value_2 ~ "Adjusted"))

flpp <- hellp %>% as_flex_table()
flpp <- set_table_properties(flpp, width = .8, layout = "autofit")

flpp <- theme_vanilla(flpp)
flpp <- set_caption(flpp, caption = "Adjusted and unadjusted factors")
flpp <- add_footer_lines(flpp, "Multivariable logistic regression analysis of the association between set variables and OFI")
```

```{r var-adjusted, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
flpp
```

As seen in Fig.\@ref(fig:care-level) the odds of opportunities for improvement differed significantly across highest hospital care level. Patients treated in a specialized ward had an OR of 15.7 (95% CI XX-XX) compared to patients treated in the ED. Patients treated in the surgical ward and intensive care unit had ORs of 10.6 (95% CI XX-XX) and 12.1 (95% CI XX-XX) respectively, compared to patients treated in the ED (Table \@ref(tab:var-adjusted)). In patients treated in a specialized ward an OFI was identified in 15.2% of the cases, compared to 1.1% in patients treated in the ED.
<!-- Det kan vara bra att presentera enhet för olika faktorer i både text och figure, främst för exv tid till dt (sekunder) osv. Det kan även vara bra att ha enhet på y-axeln på bilderna samt ev rubrik över. Sedan är vi egentligen mest intresserad av de significanta justerade variablerna varpå fokus ev bör gå dit. Men hör med Martin innan du gör massa arbete för att ändra bilderna. -->

```{r, include=FALSE}
k.sex <- aa %>% count(ofi, Gender)
k.res <- aa %>% count(ofi, res_survival)
k.intub <- aa %>% count(ofi, intub)
k.host <- aa %>% count(ofi, host_care_level)
k.rr <- aa %>% count(ofi, rr_rts)
k.sbp <- aa %>% count(ofi, sbp_rts)
k.gcs <- aa %>% count(ofi, gcs_sum)

k.sex = ddply(k.sex, .(Gender), transform, percent = n/sum(n) * 100)
k.sex = ddply(k.sex, .(Gender), transform, pos = (cumsum(n) - 0.5 * n))
k.sex$label = paste0(sprintf("%.0f", k.sex$percent))

k.res = ddply(k.res, .(res_survival), transform, percent = n/sum(n) * 100)
k.res = ddply(k.res, .(res_survival), transform, pos = (cumsum(n) - 0.5 * n))
k.res$label = paste0(sprintf("%.0f", k.res$percent))

k.intub = ddply(k.intub, .(intub), transform, percent = n/sum(n) * 100)
k.intub = ddply(k.intub, .(intub), transform, pos = (cumsum(n) - 0.5 * n))
k.intub$label = paste0(sprintf("%.0f", k.intub$percent))

k.host = ddply(k.host, .(host_care_level), transform, percent = n/sum(n) * 100)
k.host = ddply(k.host, .(host_care_level), transform, pos = (cumsum(n) - 0.5 * n))
k.host$label = paste0(sprintf("%.0f", k.host$percent))

k.rr = ddply(k.rr, .(rr_rts), transform, percent = n/sum(n) * 100)
k.rr = ddply(k.rr, .(rr_rts), transform, pos = (cumsum(n) - 0.5 * n))
k.rr$label = paste0(sprintf("%.0f", k.rr$percent))

k.sbp = ddply(k.sbp, .(sbp_rts), transform, percent = n/sum(n) * 100)
k.sbp = ddply(k.sbp, .(sbp_rts), transform, pos = (cumsum(n) - 0.5 * n))
k.sbp$label = paste0(sprintf("%.0f", k.sbp$percent))

k.gcs = ddply(k.gcs, .(gcs_sum), transform, percent = n/sum(n) * 100)
k.gcs = ddply(k.gcs, .(gcs_sum), transform, pos = (cumsum(n) - 0.5 * n))
k.gcs$label = paste0(sprintf("%.0f", k.gcs$percent))

opportunity_sex <- ggplot(k.sex, aes(x = Gender, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) + labs(y = "Percent (%)", x = "Sex") + guides(fill=guide_legend(title="OFI")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3))

opportunity_survival <- ggplot(k.res, aes(x = res_survival, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) + labs(y = "Percent (%)", x = "Survival (30 days)") + guides(fill=guide_legend(title="OFI")) + rremove("ylab") + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3))

opportunity_intub <- ggplot(k.intub, aes(x = intub, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 3) + labs(y = "Percent (%)", x = "Intubation") + rremove("ylab") + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3))

opportunity_rr <- ggplot(k.rr, aes(x = rr_rts, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 2) + labs(y = "Percent (%)", x = "Respiratory rate") + guides(fill=guide_legend(title="OFI")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3))

opportunity_sbp <- ggplot(k.sbp, aes(x = sbp_rts, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 2) + labs(y = "Percent (%)", x = "Systolic blood pressure") + guides(fill=guide_legend(title="OFI")) + rremove("ylab") + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3))

opportunity_gcs <- ggplot(k.gcs, aes(x = gcs_sum, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 2.5) + labs(y = "Percent (%)", x = "Glasgow coma scale") + guides(fill=guide_legend(title="OFI")) + rremove("ylab") + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3))

opportunity_host <- ggplot(k.host, aes(x = host_care_level, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 3) + labs(y = "Percent (%)", x = "Highest hospital care level") + guides(fill=guide_legend(title="OFI")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3))

pp<- ggarrange(opportunity_sex, opportunity_survival ,opportunity_intub, opportunity_rr, opportunity_sbp, opportunity_gcs,
          labels = c("A", "B", "C", "D","E","F"),
          ncol = 3, nrow = 2, common.legend = TRUE, legend = "bottom")

opportunity_host
pp

```

```{r, include=FALSE}
e.host<-k.host[!(k.host$ofi=="No opportunity for improvement"),]
op.host <- ggplot(e.host, aes(x = host_care_level, y = percent, fill = ofi)) +
    geom_bar(position = "dodge", stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_dodge(width = 1),
    vjust = -0.5, size = 3.5) + labs(y = "%", x = "Highest hospital care level") + guides(fill=guide_legend(title="OFI")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3, colour = "black")) + ylim(0, 20)
op.host

e.sex<-k.sex[!(k.sex$ofi=="No opportunity for improvement"),]
op.sex <- ggplot(e.sex, aes(x = Gender, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_dodge(width = 1),
    vjust = -0.5, size = 3.5) + labs(y = "%", x = "Sex") + guides(fill=guide_legend(title="OFI")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3, size=8, colour = "black")) + ylim(0, 20)
op.sex

e.res<-k.res[!(k.res$ofi=="No opportunity for improvement"),]
op.res <- ggplot(e.res, aes(x = res_survival, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_dodge(width = 1),
    vjust = -0.5, size = 3.5) + labs(y = "%", x = "Survival after 30 days") + guides(fill=guide_legend(title="OFI")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3, size=10, colour = "black")) + ylim(0, 20)
op.res

e.intub<-k.intub[!(k.intub$ofi=="No opportunity for improvement"),]
op.intub <- ggplot(e.intub, aes(x = intub, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_dodge(width = 1),
    vjust = -0.5, size = 3.5) + labs(y = "%", x = "Intubation") + guides(fill=guide_legend(title="OFI")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3, size=6, colour = "black")) + ylim(0, 20)
op.intub

e.rr<-k.rr[!(k.rr$ofi=="No opportunity for improvement"),]
op.rr <- ggplot(e.rr, aes(x = rr_rts, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_dodge(width = 1),
    vjust = -0.5, size = 3.5) + labs(y = "%", x = "Respiratory rate") + guides(fill=guide_legend(title="OFI")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3, size=10, colour = "black")) + ylim(0, 20)
op.rr

e.sbp<-k.sbp[!(k.sbp$ofi=="No opportunity for improvement"),]
e.sbp[nrow(e.sbp) + 1,] <- c("Opportunity for improvement", "1-49", 0,0,0,0)
e.sbp[nrow(e.sbp) + 1,] <- c("Opportunity for improvement", "0", 0,0,0,0)
e.sbp[nrow(e.sbp) + 1,] <- c("Opportunity for improvement", "Missing", 0,0,0,0)
e.sbp$percent<-as.numeric(e.sbp$percent)
op.sbp <- ggplot(e.sbp, aes(x = sbp_rts, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_dodge(width = 1),
    vjust = -0.5, size = 3.5) + labs(y = "%", x = "Systolic blood pressure") + guides(fill=guide_legend(title="OFI")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3, size=10, colour = "black")) + ylim(0, 20)
op.sbp

e.gcs<-k.gcs[!(k.gcs$ofi=="No opportunity for improvement"),]
op.gcs <- ggplot(e.gcs, aes(x = gcs_sum, y = percent, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_dodge(width = 1),
    vjust = -0.5, size = 3.5) + labs(y = "%", x = "Glasgow Coma Scale") + guides(fill=guide_legend(title="OFI")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.2, hjust=0.3, size=8, colour = "black")) + ylim(0, 20)
op.gcs


new.bar <- ggarrange(op.sex, op.res ,op.intub, op.rr, op.sbp, op.gcs,
          labels = c("A", "B", "C", "D","E","F"),
          ncol = 3, nrow = 2, common.legend = TRUE, legend = "none")
new.bar
```

```{r care-level, fig.cap="Opportunity for improvement presented in the factor Highest hospital care level. ED = Emergency department, Admissioned = admissioned to general ward", echo=FALSE}

op.host
```

Fig.\@ref(fig:restbarchart) displays the proportion of OFI in the remaining categorical factors. OFI was found in 5.8% of the patients with a systolic blood pressure higher than 89, and 17.6% in patients with a systolic blood pressure between 50-75. No patient with a systolic blood pressure lower than 50 had any OFI registered. Patients intubated in the hospital also had a higher percentage of OFI compared to patients not intubated or intubated before arrival at the hospital. 

```{r restbarchart, fig.cap= "Opportunity for improvement presented in the remaining factors. Intub = Intubated, Pre-intub = Patient intubated before arrival at the emergency department", echo=FALSE}

new.bar
```


```{r, include=FALSE}
ppp <- ggplot(data = aa, aes(x=ofi, y=ISS)) + 
             geom_boxplot(aes(fill=ofi))+ labs(x = "OFI") + theme(axis.text.x=element_blank(), axis.title.x = element_blank(),legend.title= element_blank())
ppp

qqq <- ggplot(data = aa, aes(x=ofi, y=pt_age_yrs)) + 
             geom_boxplot(aes(fill=ofi)) + labs(y = "Age (Years)", x = "OFI") + theme(axis.text.x=element_blank(), axis.title.x = element_blank(),legend.title= element_blank())
qqq

ccc <- ggplot(data = aa, aes(x=ofi, y=dt_ed_first_ct)) + 
             geom_boxplot(aes(fill=ofi)) + scale_y_log10() + labs(y = "Time to first CT (Sec)", x = "OFI") + theme(axis.text.x=element_blank(), axis.title.x = element_blank(),legend.title= element_blank())
ccc


new.cont <- ggarrange(ppp,qqq,ccc,
          labels = c("A", "B", "C"),
          ncol = 3, nrow = 1, common.legend = TRUE, legend = "bottom")

new.cont
```

Fig.\@ref(fig:boxchart) compares ISS, age and time to first CT between patients with and without OFI. The median age for patients with OFI is 49, which is 7 years higher than the median for the group of patients without OFI. The median ISS is also 12 units higher in the OFI-group, 17 compared to 5 in no OFI. The factor ISS had p-values lower than 0.001 in both the adjusted and unadjusted analyses. Age was also statistically significant in both analyses. The factor time to first CT was not significantly different in patients with OFI compared to those without.

```{r boxchart, fig.cap= "Box plot displaying the distrubtion of patients with OFI and those without in the continuous variables. C displays the distrubtion in regard to time to first CT in logarithmic scale", echo=FALSE}
new.cont
```


\newpage

# Discussion
We found that age, hospital care level and ISS were significantly associated with OFI in both unadjusted and adjusted analyses<!-- Correct?-->. The factor most strongly associated with OFI was the highest hospital care level, with an odds ratio of 15.7 and 12.1 in specialized wards and ICU respectively, compared to the ED. This means that patients treated in the specialized wards and ICU are more likely to be exposed to preventable errors. A previous study in the US has shown that more than half of the preventable and probably preventable deaths occur in the ICU [@Teixeira2007]. This could be explained by the complexity, severeness and amount of intervention needed for cases treated in the ICU. According to the same study the most prevalent OFI were related to airway management and perioperative care and those patients are usually found in the ICU[@Teixeira2009]. The cases in the specialized wards<!-- I think that we need to explain what we mean by specialized ward, preferably in the Setting section--> are also complex, although these wards lack the access to the same amount of resources and personnel as the ICU. This could explain why the odds ratio of OFI is even higher in specialized wards compared to the ICU. Patients with ED as the highest hospital care level are usually less injured and therefore need less intervention which in turn leads to lower odds of OFI. 

It is important to acknowledge that OFI is not the same as preventable deaths considering that OFI only imply that there is a preventable error but not the magnitude of that error. OFI does not necessarily mean that the patient die either. This study shows that there is an association between death within 30 days after trauma and OFI. Before the adjustment for other factors the p-value for survival after 30 days was not significant. There is however a significant p-value after the adjustment although it suggests that the odds of an OFI is higher in patients that are alive after 30 days. The reason behind this could be that patients that die within 30 days usually are more injured. This could mean higher ISS, higher hospital care level and intubation of the patient. After the adjustment for these and other variables the model shows lower odds of OFI in patients that die within 30 days.

OFI was identified in 6.8% of all trauma deaths at Karolinska University Hospital. A study performed by colleagues at the same hospital identified at least one preventable error in 21% of all trauma deaths, although only 4% of all trauma deaths were found to be preventable [@Ghorbani2018]. That study was published 2014 and OFI was registered through the decision made by two trauma surgeons. A possible explanation to the decreased numbers of OFI in trauma deaths could be due to the adaption of new treatment strategies and the implementation of quality improvement programs. However the criteria used to determine the presence of OFI during that time was different, and therefore the results do not necessarily mean a decrease in OFI.

The factor "Time to intervention" had to be excluded due to high missing frequency (81%). Exclusion of all patients with missing in "Time to intervention" would lead to a shift in the results as there are many categories in some factors that would be without any patients (Table \@ref(tab:sample-character)). The removal of one factor is not optimal but we have to take into account that most of the patients in the ED for instance would be excluded due to low percentage of intervention and that factor would be missing. That would in turn lead to misleading results. The decision was made to remove that factor.

Other factors associated with opportunities for improvement are the continuous factors age and ISS. As seen in Table \@ref(tab:sample-characteristics) the median age differs between the two different groups, it is higher by 7 years in the OFI group. The reason behind the higher percentage of OFI in older patients is still unclear. A systematic review published 2017 identified many studies where lower healthcare quality for elders was potentially justified by the lower chance of survival [@Salway2017]. The argument that focusing more resources on younger patients could lead to a lower mean value of disability-adjusted life years could be made. However, more research is needed to clarify if such a prejudice is present.

The median ISS was 17 in the group with OFI compared to 5 in cases with no OFI. The same arguments could be made here that OFI are more likely to be identified in patients with more complex injuries. This could be due to lack of training or protocols in the care of patients with severe injuries. Another explanation could be due to lack of resources needed for severe injuries. Two studies performed in Germany showed higher numbers of preventable errors in patients with ISS > 15 compared to the whole population. In Cologne when analyzing preventable and potentially preventable deaths in patients with ISS > 15 they estimated it to be 20.2% of all trauma deaths [@Schoeneberg2016]. When calculating the same numbers for all patients, regardless of ISS, in Berlin it was estimated to be 15.1% of all trauma deaths [@Kleber2013].

When analyzing the vital signs included, some associations with OFI was found. The model failed to estimate an association between three groups in systolic blood pressure, two groups in respiratory rate and OFI respectively. This could be due to the lack of patients in these groups. In respiratory rate >29 an OFI was found in 11.0% of the cases, compared to 5.8% in 10-29. A study performed in Iran showed similar findings when comparing preventable deaths between patients with a respiratory rate <20 and 20 or higher. It was found that deaths in patients with 20 or higher in respiratory rate were more likely to be preventable [@Davoodabadi2021]. A theory explaining this could be that hyperventilation is usually a sign of severity in trauma. The hyperventilation could be due to lactic acidosis. The lactate is a anaerobic metabolism and its accumulation in trauma patients is likely due to a severe hemorrhagic shock [@Richards2018]. Hyperventilation could also be a sign of comorbidity, a predictor of morbidity and mortality in trauma patients [@Mubang2015]. These complications could be the reason behind higher percentage of OFI in patients with respiratory rate over 29.

Statistical significance was also identified in patients intubated at the hospital, but not in those intubated before arrival. Intubation at the hospital contributed to 2.8 times higher odds of an OFI compared to non-intubated patients. As mentioned earlier, GCS less than 9 and not intubated was an audit filter at this hospital. In this study GCS was unfortunately not combined with intubation, however the reason behind higher percentage of OFI in patients intubated after the arrival at the hospital could be explained by the delay of intubation in these cases. 

The primary limitation was the removal of one factor, time to intervention, due to the fact that it had missing values in approximately 81% of the cases and the results would have been misleading since there is a correlation between less complex cases and no intervention. Another limitation was the fact that the original plan included data from 2014 and onward, however, due to complications with access to the data and heterogeneous data registration it was decided to only use data from 2017 and onward. Even though all trauma 1 patients were reviewed in one way or another, the selection for review by the morbidity and mortality professionals relied mostly on audit filters. The audit filters are as mentioned earlier not yet proven to be effective, meaning that there is a risk of selection bias.

Future studies are globally needed to achieve the attention that is required to decrease the number of OFI in trauma care. It is also needed since the quality of care can differ between hospitals. This is not only a problem in trauma care but in all specialities. As discussed earlier this study did not examine the effectiveness of audit filters per se but only factors. Audit filters such as GCS under 9 and not intubated could be examined by the association of two different factors. The same could also be done on the other audit filters using multivariable logistic regression. Other studies could also be done to further investigate the reasons behind the higher prevalence of OFI in the ICU and specialized wards.

In conclusion more severe traumas that lead to abnormal vital signs and higher hospital care level were usually found to be significantly associated with OFI after the adjustment for other factors. Some adjustments could be made to make the selection of cases for morbidity and mortality conferences more efficient to save resources. Future efforts should focus on the injustice in the quality of care for different patients in trauma. Development of models trained to identify cases with OFI is a possibility.

# Appendix

Table of sample characteristics, if Time to procedure was included (Table \@ref(tab:sample-character))

```{r sample-character, fig.align='center', echo = FALSE}
table2_t <- table1(~ Gender + pt_age_yrs + res_survival + host_care_level + intub + ISS + rr_rts + sbp_rts + gcs_sum + dt_ed_first_ct + dt_ed_emerg_proc | ofi, data=app.aa, caption = "Sample characteristics, if Time to procedure was included")

table2.conv <- t1kable(table2_t) %>%
  kable_styling(latex_options = "HOLD_position", full_width = F, position = "center")

table2.conv
```

Table with unadjusted and adjusted factors with odds ratio, confidence interval and p-value respectively (Table \@ref(tab:var-time))

```{r, include=FALSE}

tbl_merge_test3 <-
    tbl_merge(tbls = list(app.aa %>%
                              select(ofi, Gender, pt_age_yrs, res_survival, rr_rts, sbp_rts, gcs_sum, intub, host_care_level, ISS, dt_ed_first_ct, dt_ed_emerg_proc) %>%
                              tbl_uvregression(
                                  method = glm,
                                  y = ofi,
                                  method.args = list(family = binomial),
                                  exponentiate = TRUE,
                                  pvalue_fun = ~style_pvalue(.x, digits = 2),
                                  hide_n = TRUE
                              )%>% bold_p(), tbl_regression(app.outcome_log,exponentiate = TRUE,pvalue_fun = ~style_pvalue(.x, digits = 2)) %>% bold_p() %>% bold_labels()))


app.hellp<- tbl_merge_test3 %>% modify_spanning_header(list (estimate_1 ~ "Unadjusted", ci_1 ~ "Unadjusted", p.value_1 ~ "Unadjusted", estimate_2 ~ "Adjusted", ci_2 ~ "Adjusted", p.value_2 ~ "Adjusted"))



app.flpp <- app.hellp %>% as_flex_table()
app.flpp <- set_table_properties(app.flpp, width = .8, layout = "autofit")

app.flpp <- theme_vanilla(app.flpp)
app.flpp <- set_caption(app.flpp, caption = "Adjusted and unadjusted factors")
app.flpp <- add_footer_lines(app.flpp, "Unadjusted and adjusted factors with odds ratio, confidence interval and p-values")

app.flpp
```

```{r var-time, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
app.flpp
```

# References

