---
title: "Factors associated with opportunities for improvement in trauma care"
output: bookdown::pdf_document2
always_allow_html: true
bibliography: citations.bib
csl: bmcemerg.csl
link-citations: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

<style type="text/css">

h1.title {
  font-size: 38px;
  color: Black;
  text-align: center;
}

</style>


```{r library, include=FALSE}
library(readr)
library(naniar)
library(plyr)
library(dplyr)
library("questionr")
library(table1)
library(ggplot2)
library(gridExtra)
library(gtsummary)
library(DiagrammeR)
library(PRISMAstatement)
library(knitr)
library(gt)
library(tidyverse)
library(glue)
library(tinytex)
library(rmarkdown)
library(RMariaDB)
library(DBI)
library(keyring)
library(dotenv)
library(gtsummary)
library(flextable)
library(kableExtra)
library(ggpubr)
library(lubridate)
library(parameters)
```


```{r include=FALSE}
conn <- DBI::dbConnect(drv = RMariaDB::MariaDB(),
                       user = 'husseina',
                       password = Sys.getenv("DB_PASSWORD"),
                       db = "opportunities_for_improvement")


dataset.names <- setNames(nm = c("swetrau", "fmp", "atgarder", "problem"))
datasets <- lapply(dataset.names, function(dataset.name) dbReadTable(conn = conn, name = dataset.name))

attach(datasets)

swetrau$arrival <- as.POSIXct(strptime(swetrau$DateTime_ArrivalAtHospital, format = "%Y-%m-%d %H:%M"))
fmp$arrival <- as.POSIXct(strptime(fmp$Ankomst_te, format = "%Y%m%d %H:%M"))
problem$arrival <- as.POSIXct(strptime(problem$Ankomst_te, format = "%Y%m%d %H:%M"))
swetrau$id <- paste(swetrau$arrival, swetrau$PersonIdentity, swetrau$TempIdentity)
fmp$id <- paste(fmp$arrival, fmp$Personnummer, fmp$Reservnummer)
problem$id <- paste(problem$arrival, problem$Personnummer, problem$Reservnummer)

# Combine datasets
combined.data <- merge(fmp, problem, by = "id", all.x = TRUE)
combined.data <- merge(combined.data, swetrau, by = "id", all.x = TRUE)

detach(datasets)
combined.datasets<-filter(combined.data, as.Date(strptime(combined.data$Ankomst_te.x, format = "%Y%m%d %H:%M")) >="2017-01-01")

create_ofi <- function(data) {
    ## Check arguments
    assertthat::assert_that(is.data.frame(data))
    assertthat::assert_that(all(c("VK_avslutad", "Problemomrade_.FMP", "Fr1.14") %in% names(data)))
    ## Create ofi variable
    data$Problemomrade_.FMP <- tolower(data$Problemomrade_.FMP)
    levels.Problemomrade_.FMP <- unique(data$Problemomrade_.FMP)
    original.levels.Problemomrade_.FMP <- c(NA, "ok", "triage på akutmottagningen",
                                            "resurs", "lång tid till op", "lång tid till dt",
                                            "vårdnivå", "traumakriterier/styrning",
                                            "missad skada", "kommunikation", "neurokirurg",
                                            "föredömligt handlagd", "logistik/teknik",
                                            "dokumentation", "dokumetation", "bristande rutin", 
                                            "handläggning", "kompetens brist", "tertiär survey")
    if (!all(levels.Problemomrade_.FMP %in% original.levels.Problemomrade_.FMP))
        stop ("Levels in Problemomrade._FMP have changed.")
    levels.Fr1.14 <- unique(data$`Fr1.14`)
    original.levels.Fr1.14 <- c(NA, "1", "3", "2", "999")
    if (!all(levels.Fr1.14 %in% original.levels.Fr1.14))
        stop ("Levels in Fr1.14 have changed.")
    prob.filters <- with(data, `Problemomrade_.FMP` != "ok" & `Problemomrade_.FMP` != "föredömligt handlagd")
    prob.mortality <- with(data, `Fr1.14` == "2" | `Fr1.14` == "3")
    prob <- prob.filters | prob.mortality
    data$VK_avslutad <- tolower(data$VK_avslutad)
    levels.VK_avslutad <- unique(data$VK_avslutad)
    original.levels.VK_avslutad <- c("ja", NA, "nej")
    if (!all(levels.VK_avslutad %in% original.levels.VK_avslutad))
        stop ("Levels in VK_avslutad have changed.")
    quality.process.done <- data$VK_avslutad == "ja"
    ofi <- ifelse(prob, "Yes",
           ifelse(quality.process.done & !prob, "No", NA))
    ofi[quality.process.done & is.na(prob)] <- "No"
    return (ofi)
}

combined.datasets$ofi <- create_ofi(combined.datasets)
combined.datasets$ofi <- with(combined.datasets, ifelse(is.na(ofi) | ofi == "No", "No", "Yes"))


# Gör en kolumn med intub: 1, intub ED, 2. Ej intub. 3. Intub prehosp.
combined.datasets$intub <- with(combined.datasets, ifelse(`pre_intubated` == 1 & is.na(combined.datasets$pre_intubated) == FALSE, 3, `ed_intubated`))
# Förklaring: tub<- om pre inte är Na och är 1(ja), sant: konvertera till 3, Falskt: använd ed_intubated (1 är ja och 2 är nej)

# Gör om 999 till Na
ab<-combined.datasets %>% replace_with_na(replace = list( "res_survival" = 999))
ac<-ab %>% replace_with_na(replace = list( "intub" = 999))
ada<-ac %>% replace_with_na(replace = list( "host_care_level" = 999))
adab<-ada %>% replace_with_na(replace = list( "ed_sbp_rtscat" = 999))
adac<-adab %>% replace_with_na(replace = list( "ed_rr_rtscat" = 999))
adad<-adac %>% replace_with_na(replace = list( "pre_rr_rtscat" = 999))
adae<-adad %>% replace_with_na(replace = list( "ed_sbp_value" = 999))
adah<-adae %>% replace_with_na(replace = list( "ed_rr_value" = 999))
adag<-adah %>% replace_with_na(replace = list( "pre_rr_value" = 999))
adaj<-adag %>% replace_with_na(replace = list( "ed_gcs_sum" = 999))
swetrau<-adaj %>% replace_with_na(replace = list( "pre_gcs_sum" = 999))

# Gör om blodtryck till RTS
swetrau$ed_sbp_value[is.na(swetrau$ed_sbp_value)]<-350

swetrau$ed_sbp_value[swetrau$ed_sbp_value>=1 & swetrau$ed_sbp_value<=49]<-1
swetrau$ed_sbp_value[swetrau$ed_sbp_value>=50 & swetrau$ed_sbp_value<=75]<-2
swetrau$ed_sbp_value[swetrau$ed_sbp_value>=76 & swetrau$ed_sbp_value<=89]<-3
swetrau$ed_sbp_value[swetrau$ed_sbp_value>89 & swetrau$ed_sbp_value<=298]<-4
swetrau$ed_sbp_value[swetrau$ed_sbp_value==350]<-5

# Tar RTS värdet från ed, om det inte finns så använder den vanliga blodtrycket, som är grupperat enligt RTS.
swetrau$sbp_rts<-with(swetrau, ifelse(is.na(swetrau$ed_sbp_rtscat)==FALSE,ed_sbp_rtscat,ed_sbp_value))

# Samma med ed_RR
swetrau$ed_rr_value[is.na(swetrau$ed_rr_value)]<-350

swetrau$ed_rr_value[swetrau$ed_rr_value>=1 & swetrau$ed_rr_value<=5]<-1
swetrau$ed_rr_value[swetrau$ed_rr_value>=6 & swetrau$ed_rr_value<=9]<-2
swetrau$ed_rr_value[swetrau$ed_rr_value>=10 & swetrau$ed_rr_value<=29]<-4
swetrau$ed_rr_value[swetrau$ed_rr_value>29 & swetrau$ed_rr_value<=70]<-3
swetrau$ed_rr_value[swetrau$ed_rr_value==99]<-5
swetrau$ed_rr_value[swetrau$ed_rr_value==350]<-5

swetrau$ed_rr_rts<-with(swetrau, ifelse(is.na(swetrau$ed_rr_rtscat)==FALSE,ed_rr_rtscat,ed_rr_value))

# Samma med pre_RR
swetrau$pre_rr_value[swetrau$pre_rr_value>=1 & swetrau$pre_rr_value<=5]<-1
swetrau$pre_rr_value[swetrau$pre_rr_value>=6 & swetrau$pre_rr_value<=9]<-2
swetrau$pre_rr_value[swetrau$pre_rr_value>=10 & swetrau$pre_rr_value<=29]<-4
swetrau$pre_rr_value[swetrau$pre_rr_value>29 & swetrau$pre_rr_value<=70]<-3
swetrau$pre_rr_value[swetrau$pre_rr_value==99]<-5
swetrau$pre_rr_value[swetrau$pre_rr_value==350]<-5

swetrau$pre_rr_rts<-with(swetrau, ifelse(is.na(swetrau$pre_rr_rtscat)==FALSE ,pre_rr_rtscat,pre_rr_value))

swetrau$rr_rts<-with(swetrau, ifelse(intub == 3 & pre_rr_rts !=5,pre_rr_rts,ed_rr_rts))

swetrau$ed_gcs_sum[is.na(swetrau$ed_gcs_sum)]<-350

# Konvertera GCS till mild 13-15 (3), moderate 9-12 (2), severe 3-8. (1). 5 = MISSING
swetrau$ed_gcs_sum[swetrau$ed_gcs_sum>=3 & swetrau$ed_gcs_sum<=8]<-1
swetrau$ed_gcs_sum[swetrau$ed_gcs_sum>=9 & swetrau$ed_gcs_sum<=12]<-2
swetrau$ed_gcs_sum[swetrau$ed_gcs_sum>=13 & swetrau$ed_gcs_sum<=15]<-3
swetrau$ed_gcs_sum[swetrau$ed_gcs_sum==99]<-5
swetrau$ed_gcs_sum[swetrau$ed_gcs_sum==350]<-5

# Gör samma med pre_GCS
swetrau$pre_gcs_sum[is.na(swetrau$pre_gcs_sum)]<-350

swetrau$pre_gcs_sum[swetrau$pre_gcs_sum>=3 & swetrau$pre_gcs_sum<=8]<-1
swetrau$pre_gcs_sum[swetrau$pre_gcs_sum>=9 & swetrau$pre_gcs_sum<=12]<-2
swetrau$pre_gcs_sum[swetrau$pre_gcs_sum>=13 & swetrau$pre_gcs_sum<=15]<-3
swetrau$pre_gcs_sum[swetrau$pre_gcs_sum==99]<-5
swetrau$pre_gcs_sum[swetrau$pre_gcs_sum==350]<-5

# Skapa gcs_sum som blir vår nya gcs, 5=missing
swetrau$gc<-with(swetrau, ifelse(intub == 3 & pre_gcs_sum !=5,pre_gcs_sum,ed_gcs_sum))
swetrau$gcs<-with(swetrau, ifelse(intub == 1 & gc ==5,99,gc))
swetrau$gcs_sum<-with(swetrau, ifelse(intub == 3 & gcs ==5,99,gcs))

# Egen datafram för räkna missing data
fake.var<-c("Gender","res_survival","intub","host_care_level","rr_rts","sbp_rts","gcs_sum")
rak.var<-c("id","ofi","Gender","res_survival","intub","host_care_level","rr_rts","sbp_rts","gcs_sum")
cont.var<-c("ISS","dt_ed_first_ct","pt_age_yrs")
cont.procedure.var<-c("ISS","dt_ed_first_ct","pt_age_yrs","dt_ed_emerg_proc")
cat.var<-c("ofi","Gender","res_survival","intub","host_care_level","rr_rts","sbp_rts","gcs_sum")

# Gör ny dataframe för att räkna missing data
ba<-select(swetrau,rak.var,cont.var)
kl<-select(swetrau,fake.var,cont.var)
na.kl<-na.omit(kl)

# Sätt ihop databaser
calc<-ba

# Gör om till Na
calh<-calc %>% replace_with_na(replace = list('pt_age_yrs'=11))
calj<-calh %>% replace_with_na(replace = list('pt_age_yrs'=13))
calk<-calj %>% replace_with_na(replace = list('pt_age_yrs'=14))

# Gör om missing i vitalparametrar för att beräkna missing
calb<-calk %>% replace_with_na(replace = list('rr_rts'=5))
cald<-calb %>% replace_with_na(replace = list('sbp_rts'=5))
cale<-cald %>% replace_with_na(replace = list('gcs_sum'=5))

sew<-as.data.frame(freq.na(cale))

# Välj variabler
nks<-select(calk,cat.var,cont.var)

# Ta bort alla missing data
ki<- na.omit(nks)

# Kopiera dataframe
ae<-ki
aa<-ki

# Gör om till faktorer
str(aa)
aa$ofi<-as.factor(aa$ofi)
aa$Gender<-as.factor(aa$Gender)
aa$host_care_level<-as.factor(aa$host_care_level)
aa$intub<-as.factor(aa$intub)
aa$res_survival<-as.factor(aa$res_survival)
aa$rr_rts<-as.factor(aa$rr_rts)
aa$sbp_rts<-as.factor(aa$sbp_rts)
aa$gcs_sum<-as.factor(aa$gcs_sum)
aa$ISS<-as.numeric(aa$ISS)
aa$dt_ed_first_ct<-as.numeric(aa$dt_ed_first_ct)
aa$pt_age_yrs<-as.numeric(aa$pt_age_yrs)

# Ritar ut alla värden på graf
pairs(aa,col=aa$ofi)

# Resultat för varje variabel, unadjusted

sex<-glm(ofi~Gender,data = aa,family = 'binomial')
summary(sex)

surv<-glm(ofi~res_survival,data = aa,family = 'binomial')
summary(surv)

tub<-glm(ofi~intub,data = aa,family = 'binomial')
summary(tub)

care<-glm(ofi~host_care_level,data = aa,family = 'binomial')
summary(care)

resp<-glm(ofi~rr_rts,data = aa,family = 'binomial')
summary(resp)

sysbp<-glm(ofi~sbp_rts,data = aa,family = 'binomial')
summary(sysbp)

gcs_var<-glm(ofi~gcs_sum,data = aa,family = 'binomial')
summary(gcs_var)

injury<-glm(ofi~ISS,data = aa,family = 'binomial')
summary(injury)

first_ct<-glm(ofi~dt_ed_first_ct,data = aa,family = 'binomial')
summary(first_ct)


age<-glm(ofi~pt_age_yrs,data = aa,family = 'binomial')
summary(age)


# Table 1
aa$ofi <-
  factor(aa$ofi,
         levels=c("No","Yes"),
         labels=c("No opportunity for improvement",
                  "Opportunity for improvement"))

aa$Gender <-
  factor(aa$Gender, levels=c("M","K"),
         labels=c("Male",
                  "Female"))

aa$res_survival <-
  factor(aa$res_survival, levels=c("2","1"),
         labels=c("Alive",
                  "Dead"))

aa$intub <-
  factor(aa$intub, levels=c("2","3","1"),
         labels=c("Not Intubated","Pre-hospital Intubation",
                  "ED Intubation"))

aa$host_care_level <-
  factor(aa$host_care_level, levels=c("1","2","3","4","5"),
         labels=c("ED","Admissioned","surgical ward","Specialized ward","ICU"))

aa$sbp_rts <-
  factor(aa$sbp_rts, levels=c("4","3","2","1","0","5"),
         labels=c(">89","76-89","50-75","1-49","0","Missing"))

aa$gcs_sum <-
  factor(aa$gcs_sum, levels=c("3","2","1","5","99"),
         labels=c("Mild: 13-15","Moderate: 9-12","Severe: 3-8","Missing","Pre-intubated"))

aa$rr_rts <-
  factor(aa$rr_rts, levels=c("4","3","2","1","0","5"),
         labels=c("10-29",">29","6-9","1-5","0","Missing"))

label(aa$ofi)      <- "Outcome"
label(aa$ISS)      <- "Injury Severity Score"
label(aa$Gender)       <- "Sex"
label(aa$intub)     <- "Intubation"
label(aa$pt_age_yrs)       <- "Age"
label(aa$res_survival)       <- "Survival after 30 days"
label(aa$host_care_level)       <- "Highest hospital care level"
label(aa$intub)       <- "Intubation"
label(aa$rr_rts)       <- "Respiratory rate"
label(aa$sbp_rts)       <- "Systolic Blood Pressure"
label(aa$gcs_sum)       <- "GCS"
label(aa$dt_ed_first_ct)       <- "Time to first CT"

units(aa$pt_age_yrs)       <- "Years"
units(aa$dt_ed_first_ct)   <- "minutes"
units(aa$sbp_rts)       <- "RTS"
units(aa$rr_rts)       <- "RTS"

# Resultat för adjusted variabels
outcome_log<-glm(ofi~.,data = aa,family = 'binomial')
summary(outcome_log)

opo<- as.data.frame(table1(~ Gender + pt_age_yrs + res_survival + host_care_level + intub + ISS + rr_rts + sbp_rts +
         gcs_sum + dt_ed_first_ct | ofi, data=aa))

# Kod för resultat-delen
# Värden för stapel-diagram
calculation<-aa
calculation$host_care_level <-
  factor(calculation$host_care_level, levels=c("ED","Admissioned","surgical ward","Specialized ward","ICU"),
         labels=c("ED","Adm","Surg","Spec","ICU"))

calculation$intub <-
  factor(calculation$intub, levels=c("Pre-hospital Intubation","ED Intubation","Not Intubated"),
         labels=c("Pre-Intub", "Intub","Not Intub"))

calculation$gcs_sum <-
  factor(calculation$gcs_sum, levels=c("Mild: 13-15","Moderate: 9-12","Severe: 3-8","Missing","Pre-intubated"),
         labels=c("13-15","9-12","3-8","Missing","Pre-intub"))

```

```{r include=FALSE, warning=FALSE}
app.ba<-select(swetrau,rak.var,cont.procedure.var)
app.kl<-select(swetrau,fake.var,cont.procedure.var)
app.calc<-app.ba

app.calh<-app.calc %>% replace_with_na(replace = list('pt_age_yrs'=11))
app.calj<-app.calh %>% replace_with_na(replace = list('pt_age_yrs'=13))
app.calk<-app.calj %>% replace_with_na(replace = list('pt_age_yrs'=14))

# Gör om missing i vitalparametrar för att beräkna missing
app.calb<-app.calk %>% replace_with_na(replace = list('rr_rts'=5))
app.cald<-app.calb %>% replace_with_na(replace = list('sbp_rts'=5))
app.cale<-app.cald %>% replace_with_na(replace = list('gcs_sum'=5))

app.sew<-as.data.frame(freq.na(app.cale))

app.nks<-select(app.calk,cat.var,cont.procedure.var)

app.ki<- na.omit(app.nks)
app.ae<-app.ki
app.aa<-app.ki

app.aa$ofi<-as.factor(app.aa$ofi)
app.aa$Gender<-as.factor(app.aa$Gender)
app.aa$host_care_level<-as.factor(app.aa$host_care_level)
app.aa$intub<-as.factor(app.aa$intub)
app.aa$res_survival<-as.factor(app.aa$res_survival)
app.aa$rr_rts<-as.factor(app.aa$rr_rts)
app.aa$sbp_rts<-as.factor(app.aa$sbp_rts)
app.aa$gcs_sum<-as.factor(app.aa$gcs_sum)
app.aa$ISS<-as.numeric(app.aa$ISS)
app.aa$dt_ed_first_ct<-as.numeric(app.aa$dt_ed_first_ct)
app.aa$dt_ed_emerg_proc<-as.numeric(app.aa$dt_ed_emerg_proc)
app.aa$pt_age_yrs<-as.numeric(app.aa$pt_age_yrs)
str(app.aa)

app.aa$ofi <-
  factor(app.aa$ofi,
         levels=c("No","Yes"),
         labels=c("No opportunity for improvement",
                  "Opportunity for improvement"))

app.aa$Gender <-
  factor(app.aa$Gender, levels=c("M","K"),
         labels=c("Male",
                  "Female"))

app.aa$res_survival <-
  factor(app.aa$res_survival, levels=c("1","2"),
         labels=c("Dead",
                  "Alive"))

app.aa$intub <-
  factor(app.aa$intub, levels=c("3","1","2"),
         labels=c("Pre-hospital Intubation",
                  "ED Intubation","Not Intubated"))

app.aa$host_care_level <-
  factor(app.aa$host_care_level, levels=c("1","2","3","4","5"),
         labels=c("ED","Admissioned","surgical ward","Specialized ward","ICU"))

app.aa$sbp_rts <-
  factor(app.aa$sbp_rts, levels=c("4","3","2","1","0","5"),
         labels=c(">89","76-89","50-75","1-49","0","Missing"))

app.aa$gcs_sum <-
  factor(app.aa$gcs_sum, levels=c("3","2","1","5","99"),
         labels=c("Mild: 13-15","Moderate: 9-12","Severe: 3-8","Missing","Pre-intubated"))

app.aa$rr_rts <-
  factor(app.aa$rr_rts, levels=c("3","4","2","1","0","5"),
         labels=c(">29","10-29","6-9","1-5","0","Missing"))

label(app.aa$ofi)      <- "Outcome"
label(app.aa$ISS)      <- "Injury Severity Score"
label(app.aa$Gender)       <- "Sex"
label(app.aa$intub)     <- "Intubation"
label(app.aa$pt_age_yrs)       <- "Age"
label(app.aa$res_survival)       <- "Survival after 30 days"
label(app.aa$host_care_level)       <- "Highest hospital care level"
label(app.aa$intub)       <- "Intubation"
label(app.aa$rr_rts)       <- "Respiratory rate"
label(app.aa$sbp_rts)       <- "Systolic Blood Pressure"
label(app.aa$gcs_sum)       <- "GCS"
label(app.aa$dt_ed_first_ct)       <- "Time to first CT"
label(app.aa$dt_ed_emerg_proc)       <- "Time to intervention"

units(app.aa$dt_ed_emerg_proc)   <- "minutes"
units(app.aa$pt_age_yrs)       <- "Years"
units(app.aa$dt_ed_first_ct)   <- "minutes"
units(app.aa$sbp_rts)       <- "RTS"
units(app.aa$rr_rts)       <- "RTS"

app.outcome_log<-glm(ofi~.,data = app.aa,family = 'binomial')
summary(app.outcome_log)

app.log<-tbl_regression(app.outcome_log,exponentiate = TRUE)

app.log
```

# List of Abbreviations

GCS ---        Glasgow Coma Scale

ISS   ---        Injury severity score

KUH   ---        Karolinska University Hospital

SweTrau ---        Swedish National Trauma Registry

M&M   ---        Mortality and Morbidity

RTS ---        Revised Trauma Score

DALY ---       Disability-Adjusted Life Years

OFI ---       Opportunity For Improvement

ED ---       Emergency Department

ICU ---       Intensive Care Unit

# Introduction

Defined as physical injury and the body´s consequential response, trauma, is one of the leading causes of mortality and morbidity (M&M) in all age groups and the leading cause of mortality in people below the age of 44 [@David2021;@champion1990]. Approximately 4.5 million people die each year due to injury in the world [@James2020]. Motor vehicle crashes alone account for more than one million deaths and roughly between 20 and 50 million injuries annually [@Committee2013]. With an average of more than 7 days in the hospital each stay [@champion1990], trauma is one of the most prevalent reasons for admission. In 2020 road traffic accidents were third in disease burden worldwide, measured by Disability-Adjusted Life Years (DALY),  a term used to describe the impact of health problems and to measure the significance of improvement in medical care [@Murray1996;@Haagsma2015].

In many cases the outcome of the patient is mostly dependent on the quality of care acquired [@Dogrul2020], as observed in the decline of deaths by organ dysfunction, reduced from 5% of all trauma deaths to approximately 0% after the adaptation of various treatment strategies and improvement in critical care, demonstrated by a study conducted in Karolinska University Hospital (KUH) [@Ghorbani2014]. Advances in trauma care are the leading cause of the decreased number of preventable deaths according to the American College of Surgeons Committee [@Committee2013]. Inconsistencies in trauma care are prevalent in areas such as resources, diagnosis, and treatment. In a teaching hospital in Tehran, reviews of all trauma cases in 1 year displayed inappropriate care in 45% of all deaths and implied that approximately 26% of all trauma deaths were preventable [@Zafarghandi2003].Other studies have estimated preventable and potentially preventable trauma death rates up to 60% [@Konadu2020]. Airway management, inadequate chest compression, inadequate blood or fluid supply are some factors that previously were found in need of improvement in trauma care according to different studies [@Zafarghandi2003;@Maio1996].

It has for a long period of time been of great importance to investigate factors that can be improved in trauma care and different methods are used to do so. The gold standard is M&M conferences, during which specific cases are reviewed to further investigate the reason behind the mortality or morbidity [@WHO2009]. The results are used to improve the trauma care and to decrease the number of preventable errors. Although these conferences are a vital part for advancement in trauma care, it remains a process that requires a great amount of resources and is still complicated to this day.

Despite the evidence supporting the use of predefined models for case selection, there is limited data on specific factors associated with opportunities for improvement in trauma care [@Slater2020]. Therefore, methods such as audit filters are adopted in trauma quality improvement programs. Audit filters are predefined factors used in the selection of cases for review and represent an unfavorable alternation, proclaimed leading to a disadvantageous outcome [@WHO2009;@Evans2009]. Figure \@ref(fig:flowchart-process) demonstrates the selection of cases for M&M conferences using audit filters. The audit filters in KUH are systolic blood pressure under 90, Glasgow Coma Scale (GCS) less than 9 and not intubated, injury severity score (ISS) more than 15 but not admitted to the intensive care unit, time to acute intervention more than 60 minutes, time to computed tomography more than 30 minutes, and death within 30 days after trauma. A systematic review of audit filters in 2009 found no studies meeting set criteria determining the effectiveness of set audit filters [@Evans2009].

Trauma quality improvement programs rely on set filters for the selection of cases for M&M conferences. Some studies have found no major opportunities for improvement in currently used audit filters and believe further advancements are essential [@Cryer1996;@Copes1995]. We hypothesize that certain factors are associated with opportunity for improvement in trauma care. The aim of this study is to asses how patient level factors are associated with opportunity for improvement in trauma care for later development of models trained to identify cases with capacity for improvement. This can be investigated by the examination of many factors registered in our database.

```{r, include=FALSE}

DiagrammeR::grViz("
digraph graph2 {

graph [layout = dot, rankdir = RR]

# node definitions with substituted label text
node [shape = rectangle, width = 4, fillcolor = Biege]
a [label = '@@1']
b [label = '@@2']
c [label = '@@3']
d [label = '@@4']
e [label = '@@5']
f [label = '@@6']
h [label = '@@7']
g [label = '@@8']

a -> b -> c -> h -> d -> e -> f 
c->g

}

[1]: paste0('The occurence of a trauma')
[2]: paste0('Registration in SweTrau')
[3]: paste0('Primary review of audit filters by a nurse')
[4]: paste0('Secondary review by a colleague')
[5]: paste0('Multidisciplinary mortality and morbidity conference')
[6]: paste0('Consensus regarding existance of OFI')
[7]: paste0('Possible OFI identified')
[8]: paste0('No OFI identified')
") %>% DiagrammeRsvg::export_svg() %>% charToRaw() %>% rsvg::rsvg_pdf("flowchart_process.pdf")
```

```{r flowchart-process, fig.cap=" \\textbf{Flowchart describing the process of trauma cases from arrival until inclusion in Mortality and Morbidity conferences}", echo=FALSE, fig.align='center'}
knitr::include_graphics("flowchart_process.pdf")
```



# Methods

### Study design

A retrospective cohort study using data from the KUH trauma registry, and the trauma care quality database, both from SweTrau, the Swedish National Trauma Registry was conducted. Established in 2004, the trauma registry includes information of pre-hospital, hospital, and post-hospital care in accordance with the Utstein template [@Utstein2009]. Factors such as demographics, vital signs, time to procedure and time to intervention are registered [@Dick1999]. The trauma care quality database consists of the cases selected for review and the presence of opportunity for improvement is noted. The registries where linked and possible factors associated with opportunities for improvement were extracted. Bi- and multivariable logistic regression were used to determine association with the presence of opportunity for improvement.

### Setting

In Sweden, pre-hospital care is managed by paramedics and specially trained physicians. The trauma patient is triaged by EMS personnel at the scene, according to acuity. Patients are prioritized according to vital signs and mechanism of injury. For instance open head injuries are priority one. All trauma patients in Stockholm with priority one are transported to KUH to receive care by dedicated trauma teams. These teams consist of a trauma surgeon, an anaesthetist, an orthopaedic surgeon, a radiologist and specialized nurses. KUH in Solna is a level one trauma center, with direct access to radiology, operation, intensive care and interventions [@Social2015;@NKS2020].

Cases are selected for M&M conferences by specialized nurses mainly based on previously mentioned audit filters and reviewed by experienced specialists from all of the fields involved in the trauma team. Preventable errors in the care of the patient are identified in each case and registered to the database as a categorical variable. The absence of such errors is also registered. A plan to solve the problems identified is also presented by the board.

### Participants

The trauma registry includes all patients admitted with trauma team activation, regardless of ISS, as well as patients admitted without trauma team activation but found to have an ISS of more than 9. The previously mentioned audit filters are used to select patients for inclusion in the trauma care quality database and M&M review. The inclusion criteria for this study are patients included in the database, the registry and aged at least 15 years. A complete case analysis was conducted, hence patients with missing data in any of the covariates or outcome with the exception of GCS, respiratory rate, and systolic blood pressure were excluded. Data was collected for approximately 21 000 patients registered between 2017 and 2021.

```{r echo=F,results='asis',error=F,warning=F}
sewe<-filter(sew, missing != 0)

rownames(sewe)[rownames(sewe) == "res_survival"] = "Survival after 30 days"
rownames(sewe)[rownames(sewe) == "gcs_sum"] = "GCS"
rownames(sewe)[rownames(sewe) == "rr_rts"] = "Respiratory rate"
rownames(sewe)[rownames(sewe) == "intub"] = "Intubation"
rownames(sewe)[rownames(sewe) == "host_care_level"] = "Highest hospital care level"
rownames(sewe)[rownames(sewe) == "ofi"] = "Outcome"
rownames(sewe)[rownames(sewe) == "pt_age_yrs"] = "Age"
rownames(sewe)[rownames(sewe) == "dt_ed_first_ct"] = "Time to first CT"
rownames(sewe)[rownames(sewe) == "sbp_rts"] = "Systolic blood pressure"
rownames(sewe)[rownames(sewe) == "dt_ed_emerg_proc"] = "Time to intervention"
rownames(sewe)[rownames(sewe) == "Gender"] = "Sex"
```

### Variables

**Study outcome**

The study outcome is significantly associated with the outcome variable using any or all predictive variables. The outcome variable is the presence of opportunities for improvement, as labeled by the M&M review board, and defined as a binary variable with the levels "Yes - At least one opportunity for improvement identified" and "No - No opportunities for improvement identified". Presence of opportunities for improvement is defined as preventable error made in a specific patient case. Data on this outcome will be extracted from the trauma care quality database. 

**Predictors**

As our aim is to assess how patient level factors are associated with opportunity for improvement, the database used will therefore have some factors that can be further examined [@Swetrau2020]. The selection of variables to be further examined are made through discussion within the research team and partly based on locally used audit filters and factors frequently registered in the database. Selected factors to be further examined are divided into categorical and continuous variables. The categorical variables are sex, survival after 30 days, highest level of hospital care, GCS, respiratory rate, systolic blood pressure and intubation of the patient. The continuous variables are age, ISS, time from arrival at the hospital until first CT and time from arrival at the hospital until first intervention.

Systolic blood pressure and respiratory rate are registered to the database as either a continuous or a categorical value according to the Revised Trauma Score (RTS), but never both. The decision to convert the continuous values, if registered, into corresponding categorical value and treat them all as categorical was therefore made. All vital signs had to be treated similar and GCS will therefore be divided into categories based on RTS. 

### Data sources and measurements

All of the data is extracted from the KUH trauma registry and the trauma care quality database. Age and gender are registered to the registry via the patient’s personal number. Vital signs are measured on arrival to the emergency department by staff and the other variables are registered from the patient´s charts.

### Bias

Since this is not a blinded study it is of utter importance to acknowledge the risk of bias. Synthetic data will however be used to develop the analysis model and after satisfactory results implement it on the data collected from the databases. This is done to lower the risk of bias.

### Statistical methods

A complete case analysis of all patients in the Karolinska trauma care quality registry was done. Variables were handled and converted in accordance with the SweTrau Manual. Potential variables were selected based on currently used audit filters, expert consensus and common demographics. Due to GCS and respiratory rate not being registered in patients intubated before arrival at the emergency department the most reasonable solution would be to use pre-hospital values registered by paramedics, to include these patients. All statistical analysis were first done on synthetic data to ensure objectivity. Bi- and multivariable logistic regression were used to determine association with the presence of opportunity for improvement. Results are presented using a confidence level of 95% and a significance level of 95%.

# Results

The baseline characteristics of included patients are presented in Table \@ref(tab:sample-characteristics). Out of `r nrow(combined.datasets)` included in the trauma registry, between 2017 and 2021, `r nrow(combined.datasets)-nrow(na.kl)` were excluded due to missing data and age under 15, leaving a total of `r nrow(na.kl)` cases applicable for the study. The most prevalent reason for exclusion was missing data in the variable `r rownames(sewe[sewe$missing == sewe[1,1],])` with `r sewe[1,1]` patients, `r round(sewe[1,1]/nrow(combined.datasets)*100, digits = 2)`% of patients registered. The amount of cases excluded due to missing data in each variable is recorded in Table \@ref(tab:missing-data). In the trauma care quality database `r nrow(combined.datasets)-sum(is.na(combined.datasets$ofi))` of the cases were applicable for the study. After the combination of data from the trauma registry and the trauma care quality database and the removal of unqualified cases, a study group of `r nrow(aa)` is produced. Among the `r nrow(aa)` included patients, `r opo[3,4]` were male. The median age (Min, Max) for all patients was `r opo[7,4]`. The survival rate after 30 days among included patients was `r round(sum(aa$res_survival=='Alive')/nrow(aa)*100, digits = 2)`%.

```{r sample-characteristics, fig.align='center', echo = FALSE}
table1_t <- table1(~ pt_age_yrs + Gender + res_survival + host_care_level + intub + ISS + rr_rts + sbp_rts +
         gcs_sum + dt_ed_first_ct | ofi, data=aa, caption = "Sample characteristics")

table1.conv <- t1kable(table1_t) %>%
  kable_styling(latex_options = "HOLD_position", full_width = F, position = "center")

table1.conv
```

```{r missing-data, echo = FALSE, fig.align='center'}
missing_table <- knitr::kable(sewe[,1:2], format = "markdown", caption = "<strong> Missing  values in each variable</strong>")

missing_table
```

### Factors associated with preventable errors
An opportunity for improvement was identified in `r sum(aa$ofi == "Opportunity for improvement")` (`r round(sum(aa$ofi == "Opportunity for improvement")/nrow(aa)*100, digits = 2)`%) out of the cases. The unadjusted and adjusted associations of selected variables with opportunities for improvement are presented through the use of logistic regression in Table \@ref(tab:var-adjusted) with respective p-value. Statistical significance before and after adjustment for other variables was found in the variables age, ISS, GCS and highest hospital care level. In the variables respiratory rate, systolic blood pressure and intubation a significant p-value was found only before adjustment for other variables. In survival after 30 days only after adjustment for other variables.

As seen in Table \@ref(tab:sample-characteristics) the median age for patients with OFI is 49, which is 7 years higher than the median for the group of patients with no OFI. The median ISS is also 12 units higher in the OFI-group, 17 compared to 5 in no OFI. In the variable GCS, statistical significance was found in patients with missing value after adjustment. As mentioned earlier these were included in the study. Both ISS and highest hospital care level displayed p-values lower than 0.001 both adjusted and unadjusted, intubation and age only unadjusted.


```{r, include=FALSE}
tbl_merge_test <-
    tbl_merge(tbls = list(aa %>%
                              select(ofi, pt_age_yrs, Gender, res_survival, rr_rts, sbp_rts, gcs_sum, intub, host_care_level, ISS, dt_ed_first_ct) %>%
                              tbl_uvregression(
                                  method = glm,
                                  y = ofi,
                                  method.args = list(family = binomial),
                                  exponentiate = TRUE,
                                  pvalue_fun = ~style_pvalue(.x, digits = 2),
                                  hide_n = TRUE
                              )%>% bold_p(), tbl_regression(outcome_log,exponentiate = TRUE,pvalue_fun = ~style_pvalue(.x, digits = 2)) %>% bold_p() %>% bold_labels()))


hellp<- tbl_merge_test %>% modify_spanning_header(list (estimate_1 ~ "Unadjusted", ci_1 ~ "Unadjusted", p.value_1 ~ "Unadjusted", estimate_2 ~ "Adjusted", ci_2 ~ "Adjusted", p.value_2 ~ "Adjusted"))

flpp <- hellp %>% as_flex_table()
flpp <- set_table_properties(flpp, width = .8, layout = "autofit")

flpp <- theme_vanilla(flpp)
flpp <- set_caption(flpp, caption = "Adjusted and unadjusted variables")
flpp <- add_footer_lines(flpp, "Unadjusted and adjusted variables with odds ratio, confidence interval and p-values")
```

```{r var-adjusted, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
flpp
```


As seen in Figure \@ref(fig:care-level) there are clear distinctions between the different groups in hospital care level. An odds ratio of 15.7 is found in the specialized ward compared to the ED. 10.6 and 12.1 in the surgical wards and ICU respectively (Table \@ref(tab:var-adjusted)). In patients treated in a specialized ward an opportunity for improvement was identified in 15.2% of the cases, compared to 1.1% in patients treated in the ED. Results are incomplete in the variables systolic blood pressure and respiratory rate due to lack of patients in a few groups. Figure \@ref(fig:rest-barchart) displays the proportion of OFI in the remaining variables.


```{r care-level, fig.cap=" \\textbf{Opportunity for improvement presented in the variable Highest hospital care level. ED = Emergency department, Adm = administered to a normal ward, Surg = surgical ward, Spec = Specialized ward}", echo=FALSE}

# Rita ut Stapeldiagram
calculation['Percentage']='100'

ggplot(calculation, aes(fill=ofi, y=Percentage, x=host_care_level)) + geom_bar(position = "fill", stat="identity") + guides(fill=guide_legend(title="OFI")) + labs(y = "Percent (%)", x = "Survival (30 days)") + rremove("grid")
```

```{r bara testar, include=FALSE}
test.calc<- calculation
ggplot(data = test.calc, aes(x = factor(res_survival), 
                          y = prop.table(stat(count)), 
                          fill = factor(ofi), 
                          label = scales::percent(prop.table(stat(count))))) +
    geom_bar(position = "dodge") + 
    geom_text(stat = 'count',
              position = position_dodge(0.9), 
              vjust = -0.5,
              size = 3) + 
    scale_y_continuous(labels = scales::percent) + 
    labs(x = 'cyl', y = 'pct', fill = 'gear')
```



```{r, include=FALSE}

opportunity_sex <-ggplot(calculation, aes(fill=ofi, y=Percentage, x=Gender)) + geom_bar(position = "fill", stat="identity")+ rremove("ylab") + theme(legend.position="none") + rremove("grid") + labs(y = "Percent (%)", x = "Sex") + guides(fill=guide_legend(title="OFI"))




opportunity_survival <-ggplot(calculation, aes(fill=ofi, y=Percentage, x=res_survival)) + 
    geom_bar(position = "fill", stat="identity") + rremove("ylab") + theme(legend.position="none") + rremove("grid") + labs(y = "Percent (%)", x = "Survival after 30 days") + guides(fill=guide_legend(title="OFI"))




opportunity_intub <-ggplot(calculation, aes(fill=ofi, y=Percentage, x=intub)) + geom_bar(position = "fill", stat="identity")+ rremove("ylab") + theme(legend.position="none") + rremove("grid") + labs(y = "Percent (%)", x = "Intubation") + guides(fill=guide_legend(title="OFI"))




opportunity_rr <-ggplot(calculation, aes(fill=ofi, y=Percentage, x=rr_rts)) + geom_bar(position = "fill", stat="identity")+ rremove("ylab") + theme(legend.position="none") + rremove("grid") + labs(y = "Percent (%)", x = "Respiratory rate") + guides(fill=guide_legend(title="OFI"))




opportunity_sbp <-ggplot(calculation, aes(fill=ofi, y=Percentage, x=sbp_rts)) + geom_bar(position = "fill", stat="identity")+ rremove("ylab") + theme(legend.position="none") + rremove("grid") + labs(y = "Percent (%)", x = "Systolic blood pressure") + guides(fill=guide_legend(title="OFI"))




opportunity_gcs <-ggplot(calculation, aes(fill=ofi, y=Percentage, x=gcs_sum)) + geom_bar(position = "fill", stat="identity")+ rremove("ylab") + theme(legend.position="none") + rremove("grid") + labs(y = "Percent (%)", x = "Glasgow coma scale") + guides(fill=guide_legend(title="OFI"))


pp<- ggarrange(opportunity_sex, opportunity_survival ,opportunity_intub, opportunity_rr, opportunity_sbp, opportunity_gcs,
          labels = c("A", "B", "C", "D","E","F"),
          ncol = 3, nrow = 2, common.legend = TRUE, legend = "bottom")

hpp <- pp + labs(caption = "Opportunity for improvement presented in the remaining  variables. Intub = Intubated, Pre-intub = Patient intubated before arrival at the emergency department.")
```

```{r rest-barchart, echo=FALSE, fig.cap= "\\textbf{Opportunity for improvement presented in the remaining  variables. Intub = Intubated, Pre-intub = Patient intubated before arrival at the emergency department.}"}

plot(hpp)

```


# Discussion
 

# Appendix
Flowchart of excluded patients (Figure \@ref(fig:missing-fig))

```{r include=FALSE}

missing_fig <- flow_exclusions(
  incl_counts = c(nrow(combined.datasets),nrow(calk), nrow(ki)),
  total_label = "Patients in the trauma registry",
  incl_labels = c("Outcome identified", "Included in the study"),
  excl_labels = c("Outcome unidentified", "Missing data")
)

missing_fig%>% DiagrammeRsvg::export_svg() %>% charToRaw() %>% rsvg::rsvg_pdf("flowchart_ex.pdf")
```

```{r missing-fig, fig.cap=" \\textbf{Flowchart of excluded patients}", echo=FALSE, fig.align='center'}
knitr::include_graphics("flowchart_ex.pdf")
```


Table of sample characteristics, if Time to procedure was included (Table \@ref(tab:sample-character))

```{r sample-character, fig.align='center', echo = FALSE}
table2_t <- table1(~ Gender + pt_age_yrs + res_survival + host_care_level + intub + ISS + rr_rts + sbp_rts + gcs_sum + dt_ed_first_ct + dt_ed_emerg_proc | ofi, data=app.aa, caption = "Sample characteristics, if Time to procedure was included")

t1kable(table2_t) %>%
  kable_styling(latex_options = "HOLD_position", full_width = F, position = "center")
```

Table with unadjusted and adjusted variables with odds ratio, confidence interval and p-value respectively (Table \@ref(tab:var-time))

```{r, include=FALSE}

tbl_merge_test3 <-
    tbl_merge(tbls = list(app.aa %>%
                              select(ofi, Gender, pt_age_yrs, res_survival, rr_rts, sbp_rts, gcs_sum, intub, host_care_level, ISS, dt_ed_first_ct, dt_ed_emerg_proc) %>%
                              tbl_uvregression(
                                  method = glm,
                                  y = ofi,
                                  method.args = list(family = binomial),
                                  exponentiate = TRUE,
                                  pvalue_fun = ~style_pvalue(.x, digits = 2),
                                  hide_n = TRUE
                              )%>% bold_p(), tbl_regression(app.outcome_log,exponentiate = TRUE,pvalue_fun = ~style_pvalue(.x, digits = 2)) %>% bold_p() %>% bold_labels()))


app.hellp<- tbl_merge_test3 %>% modify_spanning_header(list (estimate_1 ~ "Unadjusted", ci_1 ~ "Unadjusted", p.value_1 ~ "Unadjusted", estimate_2 ~ "Adjusted", ci_2 ~ "Adjusted", p.value_2 ~ "Adjusted"))

app.flpp <- app.hellp %>% as_flex_table()

app.flpp <- theme_vanilla(app.flpp)
app.flpp <- add_footer_lines(app.flpp, "Unadjusted and adjusted variables with odds ratio, confidence interval and p-value respectively. P-values <0.05 are statistically significant and are therefore bold")
app.flpp <- set_caption(app.flpp, caption = "Adjusted and unadjusted variables")
app.flpp
```

```{r var-time, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
app.flpp
```
```{r}
heeel <- aa %>% count(ofi, Gender)

heeel = ddply(heeel, .(Gender), transform, percent = n/sum(n) * 100)
heeel = ddply(heeel, .(Gender), transform, pos = (cumsum(n) - 0.5 * n))
heeel$label = paste0(sprintf("%.0f", heeel$percent), "%")

ggplot(heeel, aes(x = factor(Gender), y = n, fill = ofi)) +
    geom_bar(position = position_stack(), stat = "identity", width = .7) +
    geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4)
```



# References

